---
title: "Pre-processing"
author: "Rochelle Terman"
date: "November 2016"
output: html_document
---

This is an **annotated version** of the following (full) lesson: https://github.com/rochelleterman/text-analysis-dhbsi/blob/master/1-Pre-processing.Rmd

### Required Pakcages

First let's load our required packages.

```{r}
setwd("~/Dropbox/berkeley/Git-Repos/PS239T/12_text-analysis")
rm(list=ls())
require(tm) # Framework for text mining
require(ggplot2) # for plotting word frequencies
require(dplyr)
require(RTextTools) # a machine learning package for text classification written in R
require(qdap) # Quantiative discourse analysis of transcripts
require(qdapDictionaries)
require(SnowballC) # for stemming
```

## 1. Prepare a Corpus

A corpus is a collection of texts, usually stored electronically, and from which we perform our analysis. A corpus might be a collection of news articles from Reuters or the published works of Shakespeare. 

Within each corpus we will have separate articles, stories, volumes, each treated as a separate entity or record. Each unit is called a "document."

Documents come in a variety of formats, but plain text (e.g. UTF-8) is best.

For this unit, we will be using a section of Machiavelli's Prince as our corpus. Since The Prince is a monograph, we have already "chunked" the text, so that each short paragraph or "chunk" is considered a "document."

### 1.1 Corpus Sources and Readers

The `tm` package supports a variety of sources and formats. 

```{r}
getSources()
getReaders()
```

Here we'll be going over two main options to input a corpus. 

Option 1) We can read a corpus from a directory that contains text files, each document a different file.
```{r}
docs <- Corpus(DirSource("Data/MachText")) 
docs
```

Option 2) We can read from a csv of documents, with each row being a document, and columns for text and metadata (information about each document). This is the easiest option if you have metadata.

```{r}
docs.df <-read.csv("Data/mach.csv", header=TRUE) #read in CSV file
docs <- Corpus(VectorSource(docs.df$text))
docs
```

Once we have the corpus, we can inspect the documents using inspect()

```{r}
# see the 16th document
inspect(docs[16])
```

And see the text using the `as.chracter`

```{r}
 # see content for 16th document
as.character(docs[[16]])
```

### 1.2 Preprocessing

Many text analysis applications follow a similar 'recipe' for preprecessing, involving:

1. Tokenizing the text to unigrams (or bigrams, or trigrams)
2. Converting all characters to lowercase
3. Removing punctuation
4. Removing numbers
5. Removing Stop Words, inclugind custom stop words
6. "Stemming" words, or lemmitization. There are several stemming alogrithms. Porter is the most popular.
7. Creating a Document-Term Matrix (DTM)

A document term matrix is simply a matrix with documents as the rows and terms as the columns and a count of the frequency of words as the cells of the matrix.

There are other, less common, preprocessing steps, which we won't be covering in this annotated unit:

8. Weighting features
9. Removing Sparse Terms

`tm` also lets us convert a corpus to a DTM while completing the pre-processing steps in one step.

```{r}
dtm <- DocumentTermMatrix(docs,
           control = list(tolower = TRUE,
                          removePunctuation = TRUE,
                          removeNumbers = TRUE,
                          stopwords = TRUE,
                          stemming=TRUE))
```

### 1.3 Exploring the DTM

Let's look at the structure of our DTM.

```{r}
# how many documents? how many terms?
dim(dtm)
# take a quick look
inspect(dtm[1:5,800:810])
```

We can obtain the term frequencies as a vector by converting the document term matrix into a matrix and summing the column counts:

```{r}
 # how many terms?
freq <- colSums(as.matrix(dtm))
length(freq)
```

By ordering the frequencies we can list the most frequent terms and the least frequent terms:

```{r}
# order
ord <- order(freq)

# Least frequent terms
freq[head(ord)]

# most frequent
freq[tail(ord)]

# frequency of frenquencies
head(table(freq),15)
tail(table(freq),15)

# plot
plot(table(freq))
```

Exploring word frequences

```{r}
# Have a look at common words
findFreqTerms(dtm, lowfreq=100) # words that appear at least 100 times

# Which words correlate with "war"?
findAssocs(dtm, "war", 0.3)

# plot the most frequent words
freq <- sort(colSums(as.matrix(dtm)),decreasing=TRUE)
head(freq)

wf <- data.frame(word=names(freq), freq=freq)
head(wf)

subset(wf, freq>50) %>%
    ggplot(aes(word, freq)) +
        geom_bar (stat ="identity") +
        theme(axis.text.x=element_text(angle=45,hjust=1))

# wordcoulds!
library(wordcloud)
set.seed(123)
wordcloud(names(freq), freq, max.words=100, colors=brewer.pal(6,"Dark2"))
```

### 1.5 Exporting the DTM

We can convert a DTM to a matrix or data.frame in order to write to a csv, add meta data, etc.

```{r}
# coerce into dataframe
dtm <- as.data.frame(as.matrix(dtm))
names(docs)  # names of documents

# add fake column for section
dtm$doc_section <- "NA"
dtm$doc_section[1:100] <- "Section 1"
dtm$doc_section[101:188] <- "Section 2"
dtm$doc_section <- as.factor(dtm$doc_section)

# check to see if they're the same number of documents per author
summary(dtm$doc_section)
````

### Execise: Make a DTM out of your own corpus or one of the toy copora.
