---
title: "Dictionary Methods and Sentiment Analysis"
author: "Rochelle Terman"
date: "November 2016"
output: html_document
---

This is an **annotated version** of the following (full) lesson: https://github.com/rochelleterman/text-analysis-dhbsi/blob/master/3-sentiment.Rmd


### Required Packages

```{r}
rm(list=ls())
setwd("~/Dropbox/berkeley/Git-Repos/PS239T/12_text-analysis")

library(ggplot2) # plotting discourse data
library(qdap) # quantiative discourse analysis of transcripts
library(data.table) # for easier data manipulation
library(scales) # to help us plot
```

The `qdap`(Rinker, 2014) package provides an extensive suite of functions to support the quantitative analysis of text. 

The package was originally intended to analyze transcript data. So the data must be in a particular format, namely a tabular spreadsheet with a `text` variable and some kind of `speaker` or metadata variable. The `text` variable must be in the form of sentences, with each sentence ending with punctuation: `.` , `?`, `!`.

`qdap` has an [extensive vignette](https://dl.dropboxusercontent.com/u/61803503/packages/qdap_vignette.html) that will help guide you through its requirements and functionality.

## 1. Comparing Songs on the Thriller Album

Let's start with a simple example using lyrics from Michael Jackson's Thriller album.

```{r}
# load data
thriller <- read.csv("Data/thriller.csv")

# see a truncated version of the data
head(truncdf(thriller),10)

# split the data into sentences (or, in this case, lines in the songs)
thril.split <- sentSplit(thriller, "Lyrics")
head(truncdf(thril.split),10)
```

The `qdap` package is great for using dictionary methods to analyze text. One of the most popular of these menthods is sentiment analysis, which calculates how "positive" or "negative" text is.

In `qdap`, we analyze sentiment using the `polarity` function.

```{r}
# calculate polarity
(poldat <- with(thril.split, polarity(Lyrics, Song)))

# have a peak of each line
counts(poldat)[1:10,]

# plot
plot(poldat)

# we can get just the second plot this way:
x <- plot(poldat)
x$p2
```

## 2. Visualizing Plot using Sentiments

`qdap` also has several datasets pre-formatted to help you explore. Let's conduct a plot analysis of Romeo and Juliet using sentiments.

```{r}
# take a look at the data
head(truncdf(rajSPLIT),10)

# put that into a data.table
raj.dat <- data.table(rajSPLIT)
```

Now that we have the data, we can add columns for cumulative word counts and polarity scores. This will allow us to capture sentiments over "time." 

```{r}
# add word counts
raj.dat[, wc := wc(dialogue,missing=0)]

# Add cumulative word count and percent complete as proxy for progression
raj.dat[, cumsum := cumsum(wc)]
raj.dat[, pct.complete := raj.dat$cumsum / sum(raj.dat$wc)]
raj.dat[, pct.complete.100 := pct.complete * 100]

# calculate polarity
poldat.raj <- with(rajSPLIT, polarity(dialogue, act, constrain = TRUE))
counts(poldat.raj)[1:10,]
polcount.raj <- na.omit(counts(poldat.raj)$polarity) # get vector of just the polarity scores

# put it into a data frame
len <- length(polcount.raj)
pol.df.raj <- data.frame(polarity = polcount.raj, Time=1:len)
```

Let's plot it.

```{r}
## Calculate background rectangles for RAJ
ends <- cumsum(rle(counts(poldat.raj)$act)$lengths)
ends
starts <- c(1, head(ends + 1, -1))
starts
rects <- data.frame(xstart = starts, xend = ends + 1,
                    Act = c("I", "II", "III", "IV", "V"))

# Plot RAJ
ggplot() + theme_bw() +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend,
                              ymin = -Inf, ymax = Inf, fill = Act), alpha = 0.17) +
  geom_line(data = pol.df.raj, aes(y=polarity, x = Time), size=1, color = "grey60") +
  geom_smooth(data = pol.df.raj, aes(y=polarity, x = Time), color="royalblue", fill="lightgray", size=1.4) +
  ylab("Average Polarity") + xlab("Duration") 


# Zoom in on that smoothed line:

ggplot() + theme_bw() +
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend,
                              ymin = -Inf, ymax = Inf, fill = Act), alpha = 0.17) +
  #geom_line(data = pol.df.raj, aes(y=polarity, x = Time), size=1, color = "grey60") +
  geom_smooth(data = pol.df.raj, aes(y=polarity, x = Time), color="royalblue", fill="lightgray", size=1.4) +
  ylab("Average Polarity") + xlab("Duration")
```

## 4. Bonus: More on QDAP

Besides polarity, `qdap` has other dictionaries to measure language formality, diversity, readability, etc. It also carries a range of tools to analyze and visualize discourse data, word usage, character usage, parts of speech, etc. 

```{r}
# gantt plots
dat <- gantt(rajSPLIT$dialogue, rajSPLIT$person)
head(dat, 12)
plot(dat)

# by grouping
dat2 <- with(rajSPLIT, gantt(dialogue, list(fam.aff, sex)))
head(dat2, 12)
plot(dat2)

# ToT plots
tot_plot(mraja1, "dialogue", grouping.var = c("fam.aff","sex"), tot = FALSE)
```