---
title: "Introduction to R"
author: "Rochelle Terman"
date: "August 12, 2015"
output: html_document
---

# Introduction to R

R is a versatile, open source programming/scripting language that's useful both for statistics but also data science. Inspired by the programming language [`S`][S].

* Open source software under [GPL][].
* Superior (if not just comparable) to commercial alternatives. R has over 5,000 user contributed packages at this time. It's widely used both in academia and industry.
* Available on all platforms.
* As a result, if you do your analysis in R, anyone can easily replicate it.
* Not just for statistics, but also general purpose programming.
* Is object oriented and functional.
* Large and growing community of peers.

[S]: http://stat.bell-labs.com/S/
[GPL]: https://www.gnu.org/licenses/gpl.html

### R Studio

There are two main ways of interacting with R: using the console or by using
script files (plain text files that contain your code).

The console window (in RStudio, the bottom left panel) is the place where R is
waiting for you to tell it what to do, and where it will show the results of a
command.  You can type commands directly into the console, but they will be
forgotten when you close the session. It is better to enter the commands in the
script editor, and save the script. This way, you have a complete record of what
you did, you can easily show others how you did it and you can do it again later
on if needed. You can copy-paste into the R console, but the Rstudio script
editor allows you to 'send' the current line or the currently selected text to
the R console using the `Ctrl-Enter` shortcut.

At some point in your analysis you may want to check the content of variable or
the structure of an object, without necessarily keep a record of it in your
script. You can type these commands directly in the console. RStudio provides
the `Ctrl-1` and `Ctrl-2` shortcuts allow you to jump between the script and the
console windows.

If R is ready to accept commands, the R console shows a `>` prompt. If it
receives a command (by typing, copy-pasting or sent from the script editor using
`Ctrl-Enter`), R will try to execute it, and when ready, show the results and
come back with a new `>`-prompt to wait for new commands.

If R is still waiting for you to enter more data because it isn't complete yet,
the console will show a `+` prompt. It means that you haven't finished entering
a complete command. This is because you have not 'closed' a parenthesis or
quotation. If you're in RStudio and this happens, click inside the console
window and press `Esc`; this should help you out of trouble.

### R Environment

**Viewing objects in your global environment and how to clean them up**

List objects in your current environment

```{r, eval = FALSE}
ls()
```

remove objects from your current environment

```{r, eval = TRUE, echo = TRUE}
x <- 5
rm(x)
#print(x)
```

remove all objects from your current environment

```{r, eval = FALSE}
rm(list = ls())
```

*Notice that we have nested one function inside another.*

**Package management**

* `install.packages("package-name")` will download a package from one of the CRAN mirrors assuming that a binary is available for your operating system. If you have not set a preferred CRAN mirror in your `options()`, then a menu will pop up asking you to choose a location.
* `library("package-name")` will load a package so you can use it. It is required at the beginning of each R session.

```{r, eval = FALSE}
install.packages(stats)
library(stats)
```

### Basic Syntax

**Comments**

Use `#` signs to comment. Comment liberally in your R scripts. Anything to the right of a `#` is ignored by R. There is no docstring, or equivalent to `"""` in R

**Assignment operator**

`<-` is the assignment operator. It assigns values on the right to objects on
the left. So, after executing `x <- 3`, the value of `x` is `3`. The arrow can
be read as 3 **goes into** `x`.  You can also use `=` for assignments but not in
all contexts so it is good practice to use `<-` for assignments. `=` should only
be used to specify the values of arguments in functions, see below.

In RStudio, typing `Alt + -` (push `Alt`, the key next to your space bar at the
same time as the `-` key) will write ` <- ` in a single keystroke.

```{r}
experiment<-"current vs. voltage"
```

**Variable Names**

Variable names can only contain letters, numbers, the underscore character, and (unlike Python) the period character. Whereas an object name like ```myobject.thing``` would point to the subclass ```thing``` of ```myobject```, R only understands the object to be freestanding.  Put another way, R interprets ```myobject.thing``` the same way that Python would interpret ```myobject_thing```.

**Printing**
In R, the contents of an object can be printed by either simply executing the the object name or calling the ```print()``` function.  Note that, unlike in Python 2, ```print()``` must have object names passed directly to it as arguments inside parentheses. (Note: This changes in Python 3)

```{r, error=FALSE}
# Executing the object name
experiment
# Printing the object R-style
print(experiment)
```

**Help**

* `?` + object opens a help page for that specific object
* `??` + object searches help pages containing the name of the object

```{r, eval = FALSE}
?mean
??mean
```
