---
title: "Subsetting"
author: "PS239T"
date: "Fall 2016"
output: html_document
---

# Subsetting

When working with data, you'll need to subset objects early and often. Luckily, R's subsetting operators are powerful and fast. Mastery of subsetting allows you to succinctly express complex operations in a way that few other languages can match. Subsetting is hard to learn because you need to master a number of interrelated concepts:

* The three subsetting operators, `[`, `[[`, and `$`.

* Important differences in behaviour for different objects (e.g., vectors, lists, factors, matrices, and data frames).

* The use of subsetting in conjunction with assignment.

This unit helps you master subsetting by starting with the simplest type of subsetting: subsetting an atomic vector with `[`. It then gradually extends your knowledge, first to more complicated data types (like dataframes and lists), and then to the other subsetting operators, `[[` and `$`. You'll then learn how subsetting and assignment can be combined to modify parts of an object, and, finally, you'll see a large number of useful applications.

### Outline

1. [Data types](#1-data-types) starts by teaching you about `[`. You'll start by learning the four ways to subset atomic vectors. You'll then learn how those four methods act when used to subset lists, matrices, and data frames.
2. [Subsetting operators](#2-subsetting-operators) expands your knowledge of subsetting operators to include `[[` and `$`, focussing on the important principles of simplifying vs. preserving.
3. In [Subsetting and assignment](#3-subassignment) you'll learn the art of subassignment, combining subsetting and assignment to modify parts  of an object.
4. [Applications](#4-applications) leads you through important, but not obvious, applications of subsetting to solve problems that you often encounter in a data analysis, using the tools above.

# 1. Data types {#1-data-types}

It's easiest to learn how subsetting works for atomic vectors, and then how it generalises to higher dimensions and other more complicated objects. We'll start with `[`, the most commonly used operator. [Subsetting operators](#subsetting-operators) will cover `[[` and `$`, the two other main subsetting operators.

### 1a. Atomic vectors

Let's explore the different types of subsetting with a simple vector, `x`. 

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
```

Note that the number after the decimal point gives the original position in the vector.

**NB:** In R, positions start at 1, unlike Python. Fun!**

There are five things that you can use to subset a vector: 

#### 1.   __Positive integers__ return elements at the specified positions:

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x
x[1]
x[c(3, 1)]

# `order(x)` gives the positions of smallest to largest values.
order(x)
x[order(x)]
x[c(1, 3, 2, 4)]

# Duplicated indices yield duplicated values
x[c(1, 1)]
```

#### 2.   __Negative integers__ omit elements at the specified positions:

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)
x[-1]
x[-c(3, 1)]
```

You can't mix positive and negative integers in a single subset:

```{r, error = TRUE}
x <- c(2.1, 4.2, 3.3, 5.4)
x[c(-1, 2)]
```

#### 3.   __Logical vectors__ select elements where the corresponding logical value is `TRUE`.

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

x[c(TRUE, TRUE, FALSE, FALSE)]
```

This is probably the most useful type of subsetting because you write the expression that creates the logical vector

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# this returns a logical vector
x > 3
x

# use a conditional statement to create an implicit logical vector
x[x > 3]
```

You can combine conditional statements with `&` (and), `|` (or), and `!` (not)

```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# combing two conditional statements with &
x > 3 & x < 5


x[x > 3 & x < 5]

# combing two conditional statements with |
x < 3 | x > 5 
x[x < 3 | x > 5]

# combining conditional statements with !
!x > 5 
x[!x > 5]
```

Another way to generate implicit conditional statements is using the `%in%` operator, which works like the `in` keywords in Python

```{r}
# generate implicit logical vectors through the %in% operator
x %in% c(3.3, 4.2)
x
x[x %in% c(3.3, 4.2)]
```

#### 4. __Character vectors__ to return elements with matching names. This only works if the vector is named.
 
```{r}
x <- c(2.1, 4.2, 3.3, 5.4)

# apply names
names(x) <- c("a", "b", "c", "d")
x

# subset using names
x[c("d", "c", "a")]

# Like integer indices, you can repeat indices
x[c("a", "a", "a")]

# Careful! names are always matched exactly
x <- c(abc = 1, def = 2)
x
x[c("a", "d")]
```

#### Exercise

Subset `country.vector` below to return every value EXCEPT "Canada" and "Brazil"

```{r eval = FALSE}
country.vector<-c("Afghanistan", "Canada", "Sierra Leone", "Denmark", "Japan", "Brazil")

# Do it using positive integers

country.vector[c(______)]

# Do it using negative integers

country.vector[-c(_____)]

# Do it using a logical vector

country.vector[c(_______)]

# Do it using a conditional statement (and an implicit logical vector)

country.vector[!________ %in% c(________)]

```

### 1b. Lists

Subsetting a list works in the same way as subsetting an atomic vector. Using `[` will always return a list; `[[` and `$`, as described below, let you pull out the components of the list.

```{r}
l <- list('a' = 1, 'b' = 2)
l
  
l[1]
l[[1]]
l['a']
```

### 1c. Matrices

The most common way of subsetting matrices (2d) is a simple generalisation of 1d subsetting: you supply a 1d index for each dimension, separated by a comma. Blank subsetting is now useful because it lets you keep all rows or all columns.

```{r}
a <- matrix(1:9, nrow = 3)
colnames(a) <- c("A", "B", "C")
a

# rows come first, then columns
a[c(1, 2), ]
a[c(T, F, T), c("B", "A")]
a[0, -2]
a[c(1,2) , -2]
```

### 1d. Data frames

Data from data frames can be addressed like matrices, using two vectors separated by a columns.

```{r}
df <- data.frame(x = 4:6, y = 3:1, z = letters[1:3])
df

# return only the rows where x == 6
df[df$x == 6, ]

# return the first and third row
df[c(1, 3), ]

# return the first and third row, and the first and second column
df[c(1, 3), c(1,2)]
```

Data frames possess the characteristics of both lists and matrices: if you subset with a single vector, they behave like lists, and return only the columns.

```{r}
# There are two ways to select columns from a data frame
# Like a list:
df[c("x", "z")]
# Like a matrix
df[, c("x", "z")]

```
But there's an important difference when you select a single column: matrix subsetting simplifies by default, list subsetting does not.

```{r}
(df["x"])
class((df["x"]))

(df[, "x"])
class((df[, "x"]))
```

See the bottom section on [Simplying and Preserving to know more](#simplify-preserve)

#### Exercises

1.  Fix each of the following common data frame subsetting errors:

```{r, eval = FALSE}
# check out what we're dealing with
mtcars

# fix
mtcars[mtcars$cyl = 4, ]
mtcars[-1:4, ]
mtcars[mtcars$cyl <= 5]
mtcars[mtcars$cyl == 4 | 6, ]

# answers
mtcars[mtcars$cyl == 4, ]
mtcars[-c(1:4), ]
mtcars[mtcars$cyl <= 5,]
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
```

2.  Why does `mtcars[1:20]` return an error? How does it differ from the 
similar `mtcars[1:20, ]`?

# 2. Subsetting operators {#2-subsetting-operators}

There are two other subsetting operators: `[[` and `$`. 

* `[[` is similar to `[`, except it can only return a single value and it allows you to pull pieces out of a list. 
* `$` is a useful shorthand for `[[` combined with character subsetting. 

### 2a. `[[`

You need `[[` when working with lists. This is because when `[` is applied to a list it always returns a list: it never gives you the contents of the list. To get the contents, you need `[[`:

>  "If list `x` is a train carrying objects, then `x[[5]]` is
> the object in car 5; `x[4:6]` is a train of cars 4-6." 
>
> --- @RLangTip

Because data frames are lists of columns, you can use `[[` to extract a column from data frames:

```{r}
mtcars

# these two are equivalent
mtcars[[1]] 
mtcars[ ,1]

# which differs from this:
mtcars[1]
```

### 2b. `$`

`$` is a shorthand operator, where `x$y` is equivalent to `x[["y", exact = FALSE]]`.  It's often used to access variables in a data frame:

```{r}
# these two are equivalent
mtcars[["cyl"]]
mtcars$cyl
```

One common mistake with `$` is to try and use it when you have the name of a column stored in a variable:

```{r}
var <- "cyl"
# Doesn't work - mtcars$var translated to mtcars[["var"]]
mtcars$var

# Instead use [[
mtcars[[var]]
```

#### Exercises

1.  Take a look at the linear model below: 

```{r}
mod <- lm(mpg ~ wt, data = mtcars)
summary(mod)
```

Extract the R squared from the model summary.

```{r}
mod.sum <- summary(mod)
mod.sum$______ # Fill in here
```

# 3. Subassignment {#3-subassignment}

All subsetting operators can be combined with assignment to modify selected values of the input vector.

```{r, error = TRUE}
x <- 1:5
x
x[c(1, 2)] <- 2:3
x

# The length of the LHS needs to match the RHS!
x[-1] <- 4:1
x

x[1] <- 4:1

# This is mostly useful when conditionally modifying vectors
df <- data.frame(a = c(1, 10, NA))
df
df$a[df$a < 5] <- 0
df
```

# 4. Applications {#4-applications}

The basic principles described above give rise to a wide variety of useful applications. Some of the most important are described below. Many of these basic techniques are wrapped up into more concise functions (e.g., `subset()`, `merge()`, `plyr::arrange()`), but it is useful to understand how they are implemented with basic subsetting. This will allow you to adapt to new situations that are not dealt with by existing functions.

### 4a. Ordering Columns

Consider we have this data frame:

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)
df
```

What if we wanted to reorder the columns so that `Region` is first? We can do so using subsetting with the names (or number) of the columns:

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)

# reorder columns using names
names(df)
df1 <- df[, c("Region", "Country", "Language")]
df1

# reorder columns using indices
names(df)
df1 <- df[, c(2,1,3)]
df1
```

One helpul function is the `order` function, which takes a vector as input and returns an integer vector describing how the subsetted vector should be ordered:

```{r}
x <- c("b", "c", "a")
order(x)
x[order(x)]
```

Knowing this, we can use `order` to reorder our columns by alphabetical order.

### 4b. Removing (or keeping) columns from data frames

There are two ways to remove columns from a data frame. You can set individual columns to `NULL`: 

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)

df$Language <- NULL
```

Or you can subset to return only the columns you want:

```{r}
df <- data.frame(
  Country = c("Iraq", "China", "Mexico", "Russia", "United Kingdom"),
  Region = c("Middle East", "Asia", "North America", "Eastern Europe", "Western Europe"),
  Language = c("Arabic", "Mandarin", "Spanish", "Russian", "English")
)

df1 <- df[, c("Country", "Region")]
df1

# using negative integers
df2 <- df[, -3]
df2
```

### 4c. Selecting rows based on a condition (logical subsetting)

Because it allows you to easily combine conditions from multiple columns, logical subsetting is probably the most commonly used technique for extracting rows out of a data frame. 

```{r}
mtcars[mtcars$gear == 5, ]
mtcars[mtcars$gear == 5 & mtcars$cyl == 4, ]
```

# 5. Bonus Items {#5-bonus-items}

### 5a. `subset()`

`subset()` is a specialised shorthand function for subsetting data frames, and saves some typing because you don't need to repeat the name of the data frame.

```{r}
subset(mtcars, gear == 5)
subset(mtcars, gear == 5 & cyl == 4)
```

### 5b. Simplifying vs. preserving subsetting {#simplify-preserve}

It's important to understand the distinction between simplifying and preserving subsetting. Simplifying subsets returns the simplest possible data structure that can represent the output, and is useful interactively because it usually gives you what you want. Preserving subsetting keeps the structure of the output the same as the input, and is generally better for programming because the result will always be the same type. Omitting `drop = FALSE` when subsetting matrices and data frames is one of the most common sources of programming errors. (It will work for your test cases, but then someone will pass in a single column data frame and it will fail in an unexpected and unclear way.)

Unfortunately, how you switch between simplifying and preserving differs for different data types, as summarised in the table below.

|             | Simplifying               | Preserving                                   |
|-------------|---------------------------|----------------------------------------------|
| Vector      | `x[[1]]`                  | `x[1]`                                       |
| List        | `x[[1]]`                  | `x[1]`                                       |
| Factor      | `x[1:4, drop = T]`        | `x[1:4]`                                     |
| Data frame  | `x[, 1]` __or__ `x[[1]]`  | `x[, 1, drop = F]` __or__ `x[1]`             |

Preserving is the same for all data types: you get the same type of output as input. Simplifying behaviour varies slightly between different data types, as described below:

*   __Atomic vector__: removes names.

```{r}
x <- c(a = 1, b = 2)
x
x[1]
x[[1]]
```

*   __List__: return the object inside the list, not a single element list.

```{r}
y <- list(a = 1, b = 2)
y
str(y[1])
str(y[[1]])
```

*   __Factor__: drops any unused levels.

```{r}
z <- factor(c("a", "b"))
z[1]
z[1, drop = TRUE]
```

*   __Matrix__: if any of the dimensions has length 1, drops that dimension.

    ```{r}
a <- matrix(1:4, nrow = 2)
a
a[1, , drop = FALSE]
a[1, ]
    ```

*   __Data frame__: if output is a single column, returns a vector instead of 
    a data frame.

    ```{r}
df <- data.frame(a = 1:2, b = 3:4)
df
str(df[1])
str(df[[1]])
str(df[, "a", drop = FALSE])
str(df[, "a"])
```