---
title: "Data Visualization"
author: "PS239T"
date: "Fall 2016"
output: slidy_presentation
---

## Atomic Types

| Type | Example |
| -----| ------- |
| numeric | `3` |
| integer | `3L` |
| character | `"3"` |
| logical | `TRUE` |


## Data Structures

|  Dimension  | Homogeneous   | Heterogeneous |
|----|---------------|---------------|
| 1d | Atomic vector | List          |
| 2d | Matrix        | Data frame    |
| nd | Array         |               |

## Types of Vectors

- character
- numeric
- logical

## Making stuff

```{r, eval = F}
# make a vector
my.vector = c(thing1, thing1, thing1) 

# add item to vector
my.vector = c(my.vector, new.item)

# make list
my.list = list(thing1, thing2, thing3)

# make matrix
my.matrix = matrix(vector, nrow = X, ncol = X)
```

## Testing and Coercion

```{r, eval = F}
# helpful inspection functions
class(obj) # returns class
str(obj) # returns sructure
length(obj) # returns lengh

# coercing atomic vectors
as.character(obj)
as.logical(obj)
as.numeric(obj)

# in general
as.thing(obj)
```

## Attributes and Names

- metadata that describes data

```{r, eval = F}
names(obj)
names(obj) < c("name1", "name2", "name3")
```

## Factors

- categorical data
- ordered or unordered

```{r, eval = F}
# basic call for unordered factor
my.factor <- factor(a.vector, levels = c("level1", "level2","level3"))

# basic call for ordered factor
my.ordered <- ordered(a.vector, levels = c("level1","level2","level3"))

# coerce something into a factor. 
# basically an abreviated version of the `factor` function.
as.factor(a.vector) 

# coerce something back into a character (or whatever) vector
as.character(my.factor)
